# Logical Operators and Control Stuctures {#control}

## Logical Operators

## Control Structures

### Selection Controls

Basically if, if/else, if/elseif/else and switch

### Iteration Controls

#### for {#forloop}

## Applications: 

### Non-parametric Bootstrap

Suppose we ask 10 students how much time they work at home for their calculus class, we obtain the following results (in hour)

```{r}
student_work = c(0, 0, 0, 0.25, 0.25, 0.75, 0.75, 1, 1.25, 4)
```

We can compute the mean time spent

```{r}
mean(student_work)
```

*ADD SOMETHING ON T TEST*

```{r}
t.test(student_work)$conf.int
```

We can see that our confidence interval includes a negative values which clearly isn't meaningful. Solution: (non-parametric) bootstrap which works as follows..... ADD SOMETHING

Here is a simple function to implement this approach:

```{r}
# Number of boostrap replications
B = 500

# Compute the length of vector
n = length(student_work)

# Confidence level
alpha = 0.05

# Initialisation of 
boot_mean = rep(NA, B)

for (i in 1:B){
  student_work_star = student_work[sample(1:n, replace = TRUE)]
  boot_mean[i] = mean(student_work_star)
}

quantile(boot_mean, c(alpha/2, 1 - alpha/2))
#hist(boot_mean, probability = TRUE)

```

