# (PART) Introduction to R {-}

# Data Structures {#data}

There are different data types that are commonly used in R among which the most important ones are the following:

- **Numeric** (or double): these are used to store real numbers. Examples: -4, 12.4532, 6.

- **Integer**: examples: 2L, 12L.

- **Logical** (or boolean): examples: `TRUE`, `FALSE`.

- **Character**: examples: `"a"`, `"Bonjour"`. 

In R there are basically five types of data structures in which elements can be stored. A data structure is said to *homogeneous* if it only contains elements of the same type (for example it only contains character or only numeric values) and *heterogenous* if it contains elements of more than one type. The five types of data structrures are commonly summarized in a table similar to the one below [see e.g. @wickham2014advanced]:

Table: (\#tab:ds) Five most common types of data structrures used in R [@wickham2014advanced].

Dimension      Homogenous    Heterogeneous
------------- ------------- ----------------
1               Vector        List
2               Matrix        Data Frame
n               Array

Consider a simple data set of the top five male single tennis players presented below:

Table: (\#tab:tennis) Five best male single tennis players as ranked by ATP  (07-15-2017).

 Name              Date of Birth         Born                Country            ATP Ranking     Prize Money    Win Percentage   Grand Slam Wins
----------------  ---------------  ---------------------- -------------------  ---------------  -------------- ---------------- ------------------
Andy Murray        15 May 1987     Glasgow, Scotland       Great Britain            1             60,449,649         78.07               9
Rafael Nadal       3 June 1986     Manacor, Spain           Spain                   2            85,920,132         82.48               15
Stan Wawrinka      28 March 1985   Lausanne, Switzerland    Switzerland             3            30,577,981         63.96              5
Novak Djokovic    22 May 1987       Belgrade, Serbia          Serbia                4            109,447,408        82.77              12
Roger Federer      8 August 1981    Basel, Switzerland      Switzerland            5            104,445,185        81.80              18

Notice that this data set contains a variety of data types; in the next sections we will use these data to illustrate how to work with five common data structures.

## Vectors 

A vector has three important properties:

```{block2, note-text, type='rmdimportant'}
- The **Type** of objects contained in the vector. The function `typeof()` returns a description of the type of objects in a vector.
- The **Length** of a vector indicates the number of elements in the vector. This information can be obtained using the function `length()`.
- **Attributes** are metadata attached to a vector. The functions `attr()` and `attributes()` can be used to store and retrive attributes (more details can be found in Section \@ref(vectattr))
```

`c()` is a generic function which combines arguments to form a vector. All arguments are coerced to a common type which is the type of the returned value, and all attributes except names are removed. For example, consider the number of grand slams won by the five players considered in the eighth column of Table \@ref(tab:tennis): 

```{r}
grand_slam_win <- c(9, 15, 5, 12, 18)
```

To display the values stored in `grand_slam_win` we could simply enter the following in the R console:

```{r}
grand_slam_win
```

Alternatively, we could have created and displayed the value by using `()` around the definition of the object itself as follows:

```{r}
(grand_slam_win <- c(9, 15, 5, 12, 18))
```

Various forms of "nested concatenation" can be used to defined vectors, for example we could also define `grand_slam_win` as

```{r}
(grand_slam_win <- c(9, c(15, 5, c(12, c(18)))))
```

This approach is often used to assemble vectors in various ways.

It is also possible to define vector with characters, for example we could define a vector with the player names as follows:

```{r}
(players <- c("Andy Murray", "Rafael Nadal", "Stan Wawrinka", 
             "Novak Djokovic", "Roger Federer"))
```

### Type 

We can evaluate the kind or type of elements that are stored in a vector using the function `typeof()`. For example, for the vectors we just created we obtain:

```{r}
typeof(grand_slam_win)
typeof(players)
```

This is a little surprising as all the elements in `grand_slam_win` are integers and it would therefore seem natural to expect this as an output of the function `typeof()`. This is because R considers any number as a "double" by default, except when adding the suffix `L` after an integer. For example:

```{r}
typeof(1)
typeof(1L)
```

Therefore, we could express `grand_slam_win` as follows:

```{r}
(grand_slam_win_int <- c(9L, 15L, 5L, 12L, 18L))
typeof(grand_slam_win_int)
```

In general, the difference between the two is relatively unimportant.

### Coercion

As indicated earlier, a vector has a homogeneous data structure meaning that it can only contain a single type among all the data types. Therefore, when more than one data type is provided, R *coerces* the data into a "shared" type. To identify this "shared" type we can use this simple rule:

\begin{equation*}
 \text{logical} < \text{integer} < \text{numeric} < \text{character},
\end{equation*}

which simply means that if a vector contains more than one data type, the "shared" type will be that of the "largest" type according to the progression shown above. For example:

```{r}
# Logical + numeric
(mix_logic_int <- c(TRUE, 12, 0.5))
typeof(mix_logic_int)

# Numeric + character
(mix_int_char <- c(14.3, "Hi"))
typeof(mix_int_char)
```


### Subsetting

Naturally, it is possible to "subset" the values of in our vector in many ways. Essentially, there are four main ways of subsetting a vector. Here we'll only discuss the first three:

- **Positive Index**: We can *access* or *subset* the $i$-th element of a vector by simply using `grand_slam_win[i]` where $i$ is a positive number between 1 and length of the vector.

```{r}
# Accesing the first element
grand_slam_win[1]

# Accesing the third and first value
grand_slam_win[c(3, 1)]

# Duplicated indices yield duplicated values
grand_slam_win[c(1, 1, 2, 2, 3, 4)]
```

- **Negative Index**: We *remove* elements in a vector using negative indices:

```{r}
# Removing the second obervation
grand_slam_win[-2]

# Removing the first and fourth obserations
grand_slam_win[c(-1, -4)]
```

- **Logical Indices**: Another useful approach is based on *logical* operators:

```{r}
# Access the first and fourth observations
grand_slam_win[c(TRUE, FALSE, FALSE, TRUE, FALSE)]
```

```{block2, type='rmdcaution'}
Note that it is not permitted to "mix" positive and negative indices. For example, `grand_slam_win[c(-1, 2)]` would produce an error message.
```

### Attributes {#vectattr}

Let's suppose that we conducted an experiment under specific conditions, say a date and a place which should be stored as attributes of the object containing the results of this experiment. Indeed, objects can have arbitrary additional attributes that are used to store metadata on the object of interest. For example:

```{r}
attr(grand_slam_win, "date") <- "07-15-2017"
attr(grand_slam_win, "type") <- "Men, Singles"
```

To display the vector with its attributes

```{r}
grand_slam_win
```

To only display the attributes we can use

```{r}
attributes(grand_slam_win)
```

It is also possible to extract a specific attribute

```{r}
attr(grand_slam_win, "date")
```

### Adding Labels {#addlab}

In some cases, it is useful to augment vector elements with labels. For example, we could define the vector `grand_slam_win` and associate the name of each corresponding athlete as a labels, i.e.

```{r}
(grand_slam_win <- c("Andy Murray" = 9, "Rafael Nadal" = 15, 
                   "Stan Wawrinka" = 5, "Novak Djokovic" = 12, 
                   "Roger Federer" = 18))
```

The main advantage of this approach is that the number of grand slams won can now be referred to by the player's name. For example:

```{r}
grand_slam_win["Andy Murray"]
grand_slam_win[c("Andy Murray","Roger Federer")]
```

All labels (athlete names in this case) can be obtained with the function `names()`, i.e.

```{r}
names(grand_slam_win)
```

### Working with Dates

When working with dates it is useful to treat them as actual dates rather than character strings that *look* like dates (to a human) but don't otherwise *behave* like dates. For example, consider a vector of three dates: `c("03-21-2015", "12-13-2011", "06-27-2008")`. 

The `sort()` function returns the elements of a vector in ascending order, but since these dates are actually just character strings that *look* like dates (to a human), R sorts them in alphanumeric order (for characters) rather than chronological order (for dates):

```{r}
# The `sort()` function sorts elements in a vector in ascending order
sort(c("03-21-2015", "12-31-2011", "06-27-2008", "01-01-2012"))
```

Converting the character strings to "yyyy-mm-dd" would solve our sorting problem, but perhaps we also want to calculate the number of days between two events that are several months or years apart. 

The `as.Date()` function is one straight-forward method for converting character strings into dates that can be used as such.  The typical syntax is of the form:

```{r, eval = FALSE}
as.Dates(<vector of dates>, format = <your format>)
```

Considering the dates of birth presented in Table \@ref(tab:tennis) we can save them in an appropriate format using:

```{r}
(players_dob <- as.Date(c("15 May 1987", "3 June 1986", "28 March 1985", 
                         "22 May 1987", "8 August 1981"), 
                       format = "%d %b %Y"))
```

Note the syntax of `format = "%d %b %Y"`.  The following table shows common format elements for use with the `as.Date()` function: 

Table: (\#tab:dateFormats) Common date formatting elements for use with `as.Date()` reproduced from [http://www.statmethods.net/](http://www.statmethods.net/input/dates.html).

Symbol   Meaning                    Example 
------- -------------------------- -------------
%d        day as a number (0-31)     01-31
%a        abbreviated weekday        Mon
%A	      unabbreviated weekday      Monday
%m        month (00-12)	             00-12
%b        abbreviated month          Jan
%B        unabbreviated month	       January
%y        2-digit year               07
%Y        4-digit year               2007


There are many advantages to recommend the `as.Date()` format (in addition to proper sorting). For example, the subtraction between two dates becomes more meaningful and yields to the difference in days between them. As an example, the number of days between Rafael Nadal's and Andy Murray's dates of birth can be obtained as

```{r}
players_dob[1] - players_dob[2]
```

In addition, subsetting becomes also more intuitive and to find out the players born after 1 January 1986 we can simply run: 

```{r}
players[players_dob > "1986-01-01"]
```

There are many other reason for using this format (or other date formats). A more detailed discussion on this topic can for example be found in [Cole Beck's notes](http://biostat.mc.vanderbilt.edu/wiki/pub/Main/ColeBeck/datestimes.pdf).

### Useful Functions with Vectors

The reason for extracting or creating vectors often lies in the need to collect information from them. For this purpose, a series of useful functions are available that allow to extract information or arrange the vector elements in a certain manner which can be of interest to the user. Among the most commonly used functions we can find the following ones 

`length()` `sum()` `mean()` `order()` and `sort()` 

whose name is self-explanatory in most cases. For example we have

```{r}
length(grand_slam_win)
sum(grand_slam_win)
mean(grand_slam_win)
```

To sort the players by number of grand slam wins, we could use the function `order()` which returns the *position* of the elements of a vector sorted in an ascending order,

```{r}
order(grand_slam_win)
```

Therefore, we can sort the players in ascending order of wins as follows

```{r}
players[order(grand_slam_win)]
```

which implies that Roger Federer won most grand slams. Another related function is `sort()` which simply sorts the elements of a vector in an ascending manner. For example,

```{r}
sort(grand_slam_win)
```

which is compact version of

```{r}
grand_slam_win[order(grand_slam_win)]
```

It is also possible to use the functions `sort()` and `order()` with characters and dates. For example, to sort the players' names alphabetically (by first name) we can use:

```{r}
sort(players)
```

Similarly, we can sort players by age (oldest first)

```{r}
players[order(players_dob)]
```

or in an reversed manner (oldest last):

```{r}
players[order(players_dob, decreasing = TRUE)]
```

There are of course many other useful functions that allow to deal with vectors which we will not mention in this section but can be found in a variety of references [see e.g. @wickham2014advanced].


### Creating sequences

When using R for statistical programming and data analysis it is very common to create sequences of numbers. Here are three common ways used to create such sequences:

- `from:to`: This method is quite intuitive and very compact. For example:

```{r}
1:3
(x <- 3:1)
(y <- -1:-4)
(z <- 1.3:3)
```

- `seq_len(n)`: This function provides a simple way to generate a sequence from 1 to an arbitrary number `n`. In general, `1:n` and `seq_len(n)` are equivalent with the notable exceptions where `n = 0` and `n < 0`. The reason for these exceptions will become clear in Section \@ref(forloop). Let's see a few examples:

```{r}
n <- 3
1:n
seq_len(n)

n <- 0
1:n
seq_len(n)
```

- `seq(a, b, by/length.out = d)`: This function can be used to create more "complex" sequences. It can either be used to create a sequence from `a` to `b` by increments of `d` (using the option `by`) or of a total length of `d` (using the option `length.out`). A few examples:

```{r}
(x <- seq(1, 2.8, by = 0.4))
(y <- seq(1, 2.8, length.out = 6))
```

This can be combined with the `rep()` function to create vectors with repeated values or sequences, for example:

```{r}
rep(c(1,2), times = 3, each = 1)
rep(c(1,2), times = 1, each = 3)
```

where the option `times` allows to specify how many times the object needs to be repeated and `each` regulates how many times each element in the object is repeated.

It is also possible to generates sequences of dates using the function `seq()`. For example, to generate a sequence of 10 dates between the dates of birth of Andy Murray and Rafael Nadal we can use

```{r}
seq(from = players_dob[1], to = players_dob[2], length.out = 10)
```

Similarly, we can create a sequence between the two dates by increment of one week (backward)

```{r}
seq(players_dob[1], players_dob[2], by = "-1 week")
```

or by increment of one month (forward)

```{r}
seq(players_dob[2], players_dob[1], by = "1 month")
```

### Example: Apple Stock Price 

Suppose that one is interested in analyzing the behavior of Apple's stock price over the last three months. The first thing that is needed to perform such analysis is to recover (automatically) today's date.  In R, this can be obtained as follows

```{r}
(today <- Sys.Date())
```

Once this is done, we can obtain the date which is exactly three months ago by using

```{r}
(three_months_ago <- seq(today, length = 2, by = "-3 months")[2])
```

With this information, we can now download Apple's stock price and represent these stocks through a candlestick chart which summarizes information on daily opening and closing prices as well as minimum and maximum prices. These charts are often used with the hope of detecting trading patterns over a given period of time.

```{r candleAAPL, message = FALSE, fig.height = 5, fig.width = 6, fig.align = "center", warning = FALSE, results = 'hide', fig.cap = "Candlestick chart for Apple's stock price for the last three months."}
library(quantmod)
getSymbols("AAPL", from = three_months_ago, to = today)
candleChart(AAPL, theme='white', type='candles')
```

Using the price contained in the object we downloaded (i.e. `AAPL`), we can compute Apple's arithmetic returns which are defined as follows

\begin{equation}
  r_t = \frac{S_t - S_{t-1}}{S_{t-1}},
\end{equation}

where $r_t$ are the returns at time *t* and $S_t$ is the stock price. This is implemented in the function `ClCl()` within the `quantmod` package. For example, we can create a vector of returns as follows 

```{r}
AAPL_returns <- na.omit(ClCl(AAPL))
```

where `na.omit()` is used to remove missing values in the stock prices vector since, if we have $n+1$ stock prices, we will only have $n$ returns and therefore the first return cannot be computed. We can now compute the mean and median of the returns over the considered period.

```{r}
mean(AAPL_returns)
median(AAPL_returns)
```

However, a statistic that is of particular interest to financial operators is the Excess Kurtosis which, for a random variable that we denote as $X$, can be defined as

\begin{equation}
  \text{Kurt} = \frac{{\mathbb{E}}\left[\left(X - \mathbb{E}[X]\right)^4\right]}{\left({\mathbb{E}}\left[\left(X - \mathbb{E}[X]\right)^2\right]\right)^2} - 3.
\end{equation}

The reason for defining this statistic as *Excess* Kurtosis lies in the fact that the standardized kurtosis is compared to that of a Gaussian distribution (whose kurtosis is equal to 3) which has exponentially decaying tails. Consequently, if the Excess Kurtosis is positive, this implies that the distribution has heavier tails than a Gaussian and therefore has higher probabilities of extreme events occurring. Given this statistic, it is useful to compute this on the observed data and for this purpose a common estimator of the excess Kurtosis is

\begin{equation}
  k = \frac{\frac{1}{n} \sum_{t = 1}^{n} \left(r_t -\bar{r}\right)^4}{\left(\frac{1}{n} \sum_{t = 1}^{n} \left(r_t -\bar{r}\right)^2 \right)^2} - 3,
\end{equation}

where $\bar{k}$ denotes the sample average of the returns, i.e.

\begin{equation}
  \bar{k} = \frac{1}{n} \sum_{i = 1}^n r_i.
\end{equation}

In R, this can simply be done as follows:

```{r}
mu <- mean(AAPL_returns)
(k <- mean((AAPL_returns - mu)^4)/(mean((AAPL_returns - mu)^2))^2 - 3)
```
 
Therefore, we observe an estimated Excess Kurtosis of `r round(k,2)` which is quite high tends to indicate the returns have a heavier tails than the normal distribution. In Chapter \@ref(#control), we will revisit this example and investigate if there is **enough evidence** to conclude that Apple's stock has Excess Kurtosis larger than zero.
 