# Control Flow {#control}

## Introduction

When you are building a larger program than the examples we considered previously, we need use various **control structures** to control the "flow" of our actions. Essentially, a control structure is "block" of code that analyzes variables and chooses a direction in which to go based on given parameters. These pieces of code represent the most basic decision-making process in computing. There exist essentially two kinds of control structures. The first ones allow to assess whether a given condition is satisfied and take appropriate in response. A simple analogy to our day to day life would be "**if** it raining outside **then** take an umbrella", we will come back to this example later in this chapter. The second kind of control structures allow to repeat of a block code multiple times. For example such approach can be used to convert a color image to gray-scale by applying the same operation(s) (i.e. same code) to each pixel of the image. In this chapter, we will first discuss the two kind control structures previously mentioned and then present various example to build our intuition.




<!-- Just like in our daily life, we need to place conditions in which we manipulate our behavior. Statistical programming also fundamentally works in the same way. We program specific behaviors for the program to follow to obtain the statistics that we may need for various tasks like regression and bootstrapping. -->





## Logical Operators

Logical operators are used very commonly across all platforms. 

| Command     | Description                |
|-------------|----------------------------|
| x `>` y     | x greater than y           |
| x `>=` y    | x greater or equals to y   |
| x `<` y     | x less than y              |
| x `<=` y    | x less than or equals to y |
| x `==` y    | x equal to y               |
| x `!=` y    | x not equal to y           |
| `!`x        | Not x                      |
| x `| `y     | x or y                     |
| x `&` y     | x and y                    |
| `isTRUE`(x) | test if all of x is TRUE   |

Logical operators can be applied to vectors as well. For example, here we create two random numerical vectors and compare them. 

```{r}
x = c(3,6,3,4,5)
y = c(2,3,4,5,6)
x>y
```

This method outputs a vector of boolean TRUE and FALSE values that perform element-wise comparisons. Note, this is called a vectorized method, which we will further mention in some detail later. `isTRUE()` checks if all the elements within the object is `TRUE`. 

```{r}
isTRUE(TRUE)
isTRUE(x>y)
```

