# R Packages



Suppose we have the following two function and that we would like to construct a package:


```{r, eval = FALSE}
mc_int = function(x_range, fun, B, seed = 1291){
  # A few checks
  # Check x_range
  if (length(x_range) != 2 || x_range[1] >= x_range[2]){
    stop("x_range is incorrectly specified")
  }

  # Check fun
  if (class(fun) != "character"){
    stop("fun is incorrectly specified and should be a character")
  }

  x = mean(x_range)
  test_fun = try(eval(parse(text = fun)), silent = TRUE)
  if (class(test_fun) == "try-error"){
    stop("fun cannot be evaluated")
  }

  # Check B
  if (B < 1){
    error("B is incorrectly specified")
  }

  # Set seed
  set.seed(seed)

  # Compute the length of the interval, i.e. (b-a)
  interval_length = diff(x_range)

  # Let's draw some uniforms to get Ui and Xi
  Ui = runif(B)
  Xi = x_range[1] + Ui*interval_length

  # Compute \hat{I}
  x = Xi
  I_hat = interval_length*mean(eval(parse(text = fun)))

  # Compute \hat{I}_2
  I2_hat = interval_length*mean((eval(parse(text = fun)))^2)
  var_I_hat = (interval_length*I2_hat - I_hat^2)/B

  # Output list
  out = list(I = I_hat, var = var_I_hat,
             fun = fun, x_range = x_range, B = B)
  class(out) = "MCI"
  out
}

plot.MCI = function(x, ...){
  obj = x
  x_range = obj$x_range
  fun = obj$fun

  Delta = diff(x_range)
  x_range_graph = c(x_range[2] - 1.15*Delta, x_range[1] + 1.15*Delta)
  x = seq(from = x_range_graph[1], to = x_range_graph[2], length.out = 10^3)
  f_x = eval(parse(text = fun))
  plot(NA, xlim = range(x), ylim = range(f_x), xlab = "x", ylab = "f(x)")
  grid()
  title(paste("Estimated integral: ", round(obj$I,4),
              " (", round(sqrt(obj$var),4),")", sep = ""))
  lines(x, f_x)
  x = seq(from = x_range[1], to = x_range[2], length.out = 10^3)
  f_x = eval(parse(text = fun))
  cols = hcl(h = seq(15, 375, length = 3), l = 65, c = 100, alpha = 0.4)[1:3]
  polygon(c(x, rev(x)), c(rep(0, length(x)), rev(f_x)),
          border = NA, col = cols[1])
  abline(v = x_range[1], lty = 2)
  abline(v = x_range[2], lty = 2)
}
```


## Step 1: Create an "empty" R package

add image part1-1 to part1-4

## Step 2: Remove "example" files

add image part2-1

Remove the following files:

- NAMESPACE
- hello.Rmd (man)
- hello.r (R)

## Step 3: Edit description file

- Replace `Authors` by `Authors@R`
- Explain roles
- Add something on liscence

## Step 4: Move your script into R file in R folder

- Group functions together per theme in different files
- Run each file seperately using `source('~/demo/R/<name of your file>.R')` and eventually debug if any error occurs.

## Step 5: Documentation

- All function you want to be able to use MUST have `#' @export`
- Here is an example:

```{r, eval = FALSE}
#' @title Simple Monte-Carlo integration
#'
#' @description Compute an approximation of the integral of the function f(x)
#' with respect to dx in the range [a, b] by Monte-Carlo integration using
#' uniform sampling.
#' @param x_range A \code{vector} of dimension 2 used to denote the integration
#' region of interest, i.e. [a, b].
#' @param fun A \code{string} containing the function to integrated. It
#' is assumed that \code{x} is used as the variable of interest.
#' @param B A \code{numeric} (integer) used to denote the number of simulation.
#' @param seed A \code{numeric} used to control the seed of the random number
#' generated by this function.
#' @return A \code{list} containing the following attributes:
#' \describe{
#'      \item{I}{Estimated value of the integral}
#'      \item{var}{Estimated variance of the estimator}
#' }
#' @author Stephane Guerrier
#' @importFrom stats runif
#' @export
#' @examples
#' mc_int(x_range = c(0,1), fun = "x^2", B = 10^5)
#' mc_int(x_range = c(0,1), fun = "x^2*sin(x^2/pi)", B = 10^5)
mc_int = function(x_range, fun, B, seed = 1291){
  # A few checks
  # Check x_range
  if (length(x_range) != 2 || x_range[1] >= x_range[2]){
    stop("x_range is incorrectly specified")
  }

  # Check fun
  if (class(fun) != "character"){
    stop("fun is incorrectly specified and should be a character")
  }

  x = mean(x_range)
  test_fun = try(eval(parse(text = fun)), silent = TRUE)
  if (class(test_fun) == "try-error"){
    stop("fun cannot be evaluated")
  }

  # Check B
  if (B < 1){
    error("B is incorrectly specified")
  }

  # Set seed
  set.seed(seed)

  # Compute the length of the interval, i.e. (b-a)
  interval_length = diff(x_range)

  # Let's draw some uniforms to get Ui and Xi
  Ui = runif(B)
  Xi = x_range[1] + Ui*interval_length

  # Compute \hat{I}
  x = Xi
  I_hat = interval_length*mean(eval(parse(text = fun)))

  # Compute \hat{I}_2
  I2_hat = interval_length*mean((eval(parse(text = fun)))^2)
  var_I_hat = (interval_length*I2_hat - I_hat^2)/B

  # Output list
  out = list(I = I_hat, var = var_I_hat,
             fun = fun, x_range = x_range, B = B)
  class(out) = "MCI"
  out
}
```

- Build documentation : `devtools::document()`
- If everyhting is OK you should have one Rd document for everything function: part5-1


# Step 6: Test your package

- Build your package
- Test if it works



