# High performance computing

## Rcpp: using C++ within R
Whereas R is a relatively new language targetting essentially statistician and data
analyst, C++ is a well-established general-purpose programming language.
This chapter does not aim at covering programmation in C++ per se, but rather
to address specific problems related to R.
The point of view taken here is that of the R programmer having no notions of C++
or similar language. We do not describe accurately notions of C++ but rather give
some necessary concepts for starting programming.

R is an [interpreted language](https://en.wikipedia.org/wiki/Interpreted_language)
which makes it particularly slow for repeating operations such as for and while loops,
or recursive calling of functions. On the other hand, C++ is a 
[compiled language](https://en.wikipedia.org/wiki/Compiled_language), it means that
before being able to use your code, it is translated (compiled) to lower-level
machine language which makes the code running very fast.

Note that here we jump straight to the C++ solution, but there are many ways to enhance 
your R code such as [vectorization](http://www.dartistics.com/fast-r-code.html) and
[memory management](http://adv-r.had.co.nz/memory.html). [Profiling](http://adv-r.had.co.nz/Profiling.html)
also helps to identify which parts of the code is the slowest.

Note eventually that C++ compiles *ahead of time* (before using the code), but other
approaches exist such as the [Julia language](https://julialang.org/) that compiles 
*just in time* (at runtime).

### Installation
Working with C++ in R does not come "out of the box", but do not worry, it is 
now a well handled procedure. You need essentially two elements: a C++ compiler and a
program that permits to connect R and C++. For the latter, you need to install 
the `Rcpp` package:

```{r,eval=F}
# For CRAN official release
install.packages("Rcpp")

# For development version
devtools::install_github("RcppCore/Rcpp")
```

For the former, it depends on your OS:   
1. On Windows, install `RTools`.   
2. On Mac, install `Xcode` from the App Store.   
3. On Linux, you need to install `r-bas-dev` (or similar) from
your package manager.

Verify that your installation works by running:
```{r, echo=TRUE, message=FALSE, warning=FALSE, include=TRUE, eval=FALSE}
require(Rcpp)
cppFunction("double sumC(double x, double y){
  double s;
  s = x + y;
  return s;
} ")
sumC(3.5, 6.5)
```

```{block2, type="rmdcaution"}
In C++, a line of code must end with a semicolumn `;`. Otherwise, compiler returns an error.
```

The function `sumC(x,y)` takes two doubles (numeric) values, `x` and `y`,
and it returns a double, the sum of the two values. Can you predict what happens if you run
the following lines? Can you explain why?

```{r, eval = FALSE}
sumC(3L, 6L)
sumC(TRUE, FALSE)
sumC(c(3,4), c(5,6))
```

### A first program
The above `cppFunction()` creates an *inline* C++ function (within your R code).
However, it is better practice to create a separate file dedicated to C++ functions.
It is possible then to compile your C++ functions and call them within R.

In RStudio, you can create a C++ file by clicking on *File* > *New File* > *C++ File*.
RStudio provides you with an example. At the top of C++ file, you should see:

```{Rcpp, eval = FALSE}
#include <Rcpp.h>
using namespace Rcpp;
```

The first line tells your program that you can access functionalities from the `Rcpp` library.
The second line tells your program that you can access those functionalities without specifying 
the `Rcpp` library. For example, suppose there is function `mean()` in `Rcpp`.
With the first line, you could use this function only by typing `Rcpp::mean()`.
Adding the second line allow you to write directly `mean()` in your program.
This is somehow equivalent as using `library(Rcpp)` in your R code.

Once the C++ code is written, it should be saved with the appropriate extension, use
`*.cc` or `*.cpp`. Now you can compile your first C++ program. Type in your R file:
```{r, eval=FALSE}
sourceCpp("file_name.cpp")
```
Try to write the `sumC()` function within a separate C++ file and compile it in R.

By default, C++ functions are not accessible within your R program.
One key element is to use the attribute `[[Rcpp::export]]` in comment
**just before the function definition**. For example:

```{Rcpp, eval=FALSE}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
double sumC(
    double x,
    double y
) {
  double s;
  s = x + y;
  return s;
}
```

```{block2, type="rmdnote"}
In C++, the symbol `//` is used for single line comment. The symbol `/*` starts a comment
on possibly multiple lines and `*/` ends the comment.
```


### Data structures
One difference you can immediately notice while working with C++ is 
that you need to tell the program what type of data structure you
intend to use, before using it. This makes the code safer, 
the programmer needs to think more deeply on the implementation details 
and it prevents from using the code in a way it was not intended. 
On the other hand, the programmer will spend (at least at the beginning)
more time conceiving the code.

Here is a list of primitive data types with their keywords:   
- `int` (4) : integer,   
- `float` (4) : single precision floating number (numeric),   
- `double` (8) : double precision floating number (numeric),   
- `bool` : logical (true or false),   
- `char` : character,   
- `void` : without any value.  

The numbers in parenthesis represents typical memory in bytes required to store the variable. 
Single and double precision floating number represents how the computer
stores real numbers in the memory. `float` requires less memory than `double`
but is less precise. As a general rule, we use only `double` as it is already the case 
in R:   

> R has no single precision data type. All real numbers are stored in double precision format.
>
> `help(double)` in R

These data types can be modified by using one (or more) of the following keywords:   
- `long` : allocate more space in memory,   
- `short` : allocate less space in memory,   
- `unsigned` : numbers take only positive value.  

For more complex data structures such as vectors and matrices, the story is slightly more complicated.
Hopefully for R programmers, `Rcpp` provides a lot of predefined structures which makes
the transition to C++ easier. Here is a non-exhaustive list:

Table : Common data structure in R with Rcpp equivalent.   

Rcpp | R | Role 
--- | --- | ---   
`NumericVector` | `c(...)` or `vector(mode = "double", ...)` | A vector of double   
`IntegerVector` | `c(...)` or `vector(mode = "integer", ...)` | A vector of integer   
`LogicalVector` | `c(...)` or `vector(mode = "bool", ...)` | A vector of boolean   
`NumericMatrix` | `matrix()` | A matrix of double   
`IntegerMatrix` | `matrix()` | A matrix of integer   
`LogicalMatrix` | `matrix()` | A matrix of boolean   
`List`| `list` | A list   
`DataFrame` | `data.frame()` | A data frame   

#### Vectors

Before using a vector in C++, you need to create one. 
For example,
```{Rcpp, eval=FALSE}
NumericVector v(10);
```
creates an empty vector of double of length 10.

To access elements of a vector, the syntax is the same as in R with one notable exception:
**accessor starts at 0**! So if you want to access the first element of a vector `v`,
you should write `v[0]`, the ith element, `v[i-1]`. Note that you can equivalently
write `v(0)` and `v(i-1)`.

Eventually, the length of the vector `v` can be obtained as follows
```{Rcpp, eval=FALSE}
int n = v.length();
```
Here the length of the vector is saved into an integer `n`.

Note that `length()` is a *method* for the *class* `NumericVector`
and uses the syntax ''dot-function''. Theses concepts are beyond
the scope of this introduction, the only to remark is that the above
notation gives the same result as if you wrote `length(v)` in R.

#### Matrices

Creating a matrix is similar to vector.
```{Rcpp, eval=FALSE}
NumericMatrix m(3,4);
```
creates an empty matrix with 3 rows and 4 columns.
This is equivalent as writing in R
```{r, eval=FALSE}
m <- matrix(nr=3,nc=4)
```

The syntax for accessing matrix elements is slightly different.
For example, accessing a column or a row can be done as follows
```{Rcpp, eval=FALSE}
// Copy first column into a vector
NumericVector v = m( _ , 0);

// Copy first row into a vector
NumericVector v = m(0, _);

// Access the ith row, jth column
double x = m(i-1, j-1); // i and j are positive integers already defined
```

The number of column/row of a matrix can be retrieved using
```{Rcpp, eval=FALSE}
// number of rows
int nr = m.nrow();

// number of columns
int nc = m.ncol();
```

#### List
List creation can be accomplished as follows:
```{Rcpp, eval=FALSE}
// Create a list from elements e1 and e2
List L = List::create(e1, e2);

// Create a list from elements e1 and e2 with names
List L = List::create(Named("name1") e1, Named("name2") e2);
```


### Control structures

Logical operators in R and C++ are similar.
Here is a list where `x` and `y` are scalar (not vectors!).

Table: Common logical operators for C++.

Operator | Description      
--- | ---   
`x > y` | `x` greater than `y`  
`x >= y` | `x` greater than or equals to `y`  
`x < y` | `x` less than `y`   
`x <= y` | `x` less than or equals to `y`   
`x == y` | `x` equals to `y`   
`x != y` | `x` not equal to `y`   
`x && y` | `x` and `y`   
`x || y` | `x` or `y`  
`!x` | not `x`   


#### if/else/elseif statements
R and C++ share the same syntax for `if`/`else`/`elseif` statements
at the exception that `else if` has a space. For example,
the following expression is both valid within R and C++:
```{r,eval=FALSE}
if (condition){
  plan A;
}else{
  plan B;
}
```

#### loops
For loops are much faster in C++ than R. 
The syntax is slightly different. 
For example, the following R for loop
```{r, eval=FALSE}
for(i in 1:10){
  print("The value is: ", i)
}
```

can be written in C++ with the following code
```{Rcpp, eval=FALSE}
for(int i(1); i<=10; i++){
  Rcpp::Rcout << "The value is: " << i << std::endl;
}
```
Three elements need to be defined:  
1. The initialization statement: an iterator (`i` here), generally of type integer, needs to be initialized 
(here `i` is initialized to 1).   
2. The test expression: the expression is evaluated and should return a boolean. If `false` the loop stops and
if `true`, the body of the for loop is evaluated.   
3. The update statement: it defines the increment of the iterator, usually set to `i++`.

Note that there exists also the decrement operator `i--`.

C++ also offers variants of assignment operator.

Table: Assignment operator in C++. `x` and `y` are already defined.   

| Operator     | Equivalence  | Description                                                     |
| ------------ | ------------ | --------------------------------------------------------------- |   
| `x = y;`     |              | Assign the value of `y` into `x`                                |
| `x += y;`    | `x = x + y;` | Take the value of `x` and adds `y`, then assign to `x`          |
| `x -= y;`    | `x = x - y;` | Take the value of `x` and substracts `y`, then assign to `x`    |      
| `x *= y;`    | `x = x * y;` | Take the value of `x` and multiplies by `y`, then assign to `x` |           
| `x /= y;`    | `x = x / y;` | Take the value of `x` and divides by `y`, then assign to `x`    |     

Therefore, `i++` is equivalent to `i+=1`.

Write a loop that takes an increment of 2, 3, ...

C++ while loop has the same syntax as R. The following expression
```{Rcpp,eval=FALSE}
while( condition ) {
  body;
}
```
is licit in both languages.

Note that C++ is also equiped with a `do ... while` loop.
This maybe useful for example in the situation where
you need a first run of the body prior to checking the condition.
```{Rcpp, eval=FALSE}
do{
  body;
}
while(condition);
```

### Functions
Creating functions in C++ is fundamentally not much different than R.
Neverthelss, C++ requires an additional layer of precision.
A function can follow this pattern:
```{Rcpp, eval=FALSE}
type function_name(type arg1, ..., type argn){
  body of the function;
  
  return obj;
}
```
First, the type of return of the function should be specified.
The `obj` that is returned should be of the same type as the one specified
at the function declaration. Each argument should have a type.
Otherwise, as in R, the body of the function contains statements that defines what the function does.

As already said, by default C++ function cannot be used with R.
One key element is to use the attribute `[[Rcpp::export]]` in comment
**just before the function definition**. For example
```{Rcpp, eval=FALSE}
// [[Rcpp::export]]
type function_name(type arg1, ..., type argn){
  body of the function;
  
  return obj;
}
```
will make `function_name` available within R.

### Comparing with R
Puting efforts in writing C++ code instead of R should be rewarding.
One way to measure this reward is by comparing the computational efficiency
of C++ function in comparison to the equivalent R function.
This can be achieved for example using the `microbenchmark` package.

For example, suppose you create a function that computes the factorial number of
an input positive integer.
In R, you can write
```{r}
my_factorialR <- function(n){
  factorial <- 1
  for(i in 2:n){
    factorial <- factorial * i
  }
  return(factorial)
}
```

An equivalent C++ function could be
```{Rcpp, eval=FALSE}
int my_factorialC(int n){
  int factorial = 1;
  for(int i(2); i<=n; ++i){
    factorial *= i;
  }
  return factorial;
}
```

```{r, cache=FALSE, echo=FALSE,include=FALSE,eval=TRUE}
Rcpp::cppFunction(
  "int my_factorialC(int n){
  int factorial = 1;
  for(int i(2); i<=n; ++i){
    factorial *= i;
  }
  return factorial;
}")
microbenchmark::microbenchmark(my_factorial(500), my_factorialC(500))
```

```{r, cache = FALSE, eval=TRUE}
require(microbenchmark)
microbenchmark(my_factorial(500), my_factorialC(500))
```


### Example: Buffon's needle (con't)
Let's reconsider Buffon's needle problem.
We propose to recode the R functions `cast_needle()` and
`buffon_experiment()` into C++.

For this purpose, first create a C++ file with header
```{Rcpp, eval=FALSE}
#include <Rcpp.h>
using namespace Rcpp;
```

The `cast_needle()` function was defined as
```{r, eval=FALSE}
cast_needle <- function(plane_width = 20){
  available_range <- plane_width/2 - 1 # where 1 is the length of the needle (unit)
  x_start <- runif(2, -available_range, available_range)
  angle <- runif(1, 0, 2*pi)
  x_end <- c(cos(angle), sin(angle)) + x_start # where the angles are multiplied by the needle length which is 1 in this example
  cross <- floor(x_start[2]) != floor(x_end[2])
  out <- list(start = x_start, end = x_end, cross = cross)
  out
}
```

A possible C++ implementation is 
```{Rcpp, eval=FALSE}
// [[Rcpp::export]]
List cast_needle(
  unsigned int plane_width = 20
){
  // Variable declaration
  int available_range;
  double angle,my_pi;
  NumericVector x_start(2), x_end(2), tmp_angle(1);
  bool cross;
  
  // Computation
  my_pi = atan(1) * 4; // defines pi (for simplicity)
  available_range = plane_width / 2 - 1;
  x_start = runif(2, -available_range, available_range);
  tmp_angle = runif(1, 0, 2 * my_pi);
  angle = tmp_angle[0];
  x_end[0] = cos(angle) + x_start[0];
  x_end[1] = sin(angle) + x_start[1];
  cross = floor(x_start[2]) != floor(x_end[2]);
  
  return List::create(
    Named("start") = x_start,
    Named("end") = x_end,
    Named("cross") = cross
  );
}
```

We have few remarks:   

- For better readibility, it is better practice to separate variable declaration and computation.  
- The value of $\pi$ is not accessible by default, here we simply approximate it but better solutions exist.   
- Basic functions such as `atan()`, `runif()`, `floor()`, `cos()`, `sin()` are made available
by `Rcpp` (check for **Rcpp sugar**).   
- To avoid an error we have first to save the result of `runif()` as `NumericVector`, 
its default result's type.   


The `buffon_needle()` function was declared in R as follows:
```{r,eval=FALSE}
buffon_experiment <- function(B = 2084, plane_width = 10, seed = NULL){
  
  if (!is.null(seed)){
    set.seed(seed)
  }
  
  X_start <- X_end <- matrix(NA, B, 2) 
  cross <- rep(NA, B)
  
  for (i in 1:B){
    inter <- cast_needle(plane_width = plane_width)
    X_start[i, ] <- inter$start
    X_end[i, ] <- inter$end
    cross[i] <- inter$cross
  }
  
  out <- list(start = X_start, end = X_end, cross = cross, plane = plane_width)
  class(out) <- "buffon_experiment"
  out
}
```

The C++ alternative we propose is the following
```{Rcpp, eval=FALSE}
// [[Rcpp::export]]
List buffon_experiment(
  int B = 2084,
  int plane_width = 10
){
  // Variable declaration
  NumericMatrix X_start(B,2), X_end(B,2);
  LogicalVector cross(B);
  List L;
  NumericVector tmp(2);
  
  // Computation
  for(int i(0); i<B; ++i){
    L = cast_needle(plane_width);
    tmp = L["start"];
    X_start(i,_) = tmp; 
    tmp = L["end"];
    X_end(i,_) = tmp;
    cross(i) = L["cross"];
  }
  
  return List::create(
    Named("start") = X_start,
    Named("end") = X_end,
    Named("cross") = cross,
    Named("plane") = plane_width
  );
}
```

We have few remarks:   

- C++ does not know the content of `L["start"]` in advanced, so 
it is saved first into a vector `tmp` of the right dimension. Same for `L["end"]`.   
- To avoid complexity, the list returned by the C++ implementation has
not the class *buffon_experiment*. This should be added within R.   
