# Basic Probability and Statistics 

## Probability Distributions 

For probability distributions, we can compute basic values such as density, distribution, quantiles, random sampling using `R` built-in functions below. 

-`dname` calculates density (pdf) at input `x`.
-`pname` calculates distribution (cdf) at input `x`.
-`qname` calculates the quantile at an input probability.
-`rname` generates a random draw from a particular distribution.

Note that we will replace `name` with the name of a given distribution. For example, a a random number generator for uniform distribution will be `runif`. To learn more about specific parameters and documentation for a function, type `?` before the function name into your `R` console. Common example probability distributions include, but are not limited to, binomial, chi-square, exponential, F, geometric, logistic, normal, poisson, t, and uniform. 

Here's an example. 

### Problem {-}

Assume that the test scores of a college entrance exam fits a normal distribution. Furthermore, the mean test score is 70, and the standard deviation is 15. What is the percentage of students scoring 90 or more in the exam?

### Answer {-}

In this case, we consider a random variable $X$ that is normally distributed $X \sim N(\mu=70, \sigma^2=25)$. Since we are looking for the percentage of students scoring higher than 90, we are interested in the upper tail of the normal distribution.

```{r}
pnorm(q = 90, mean = 70, sd = 15, lower.tail = FALSE) 
```

We see that there is a 9% probability of students scoring 90 or more in the exam. Note that at `?pnorm` we see an additional parameter, `lower.tail` that indicates TRUE (default) if probabilities are P[X ≤ x] and FALSE otherwise, where P[X > x].

## Summary Statistics 

### Numeric Input 

For central tendency or spread statistics of numeric input, we can use the following `R` build-in functions below. 

- `mean` calculates the mean of an input `x`.
- `median` calculates the median of an input `x`.
- `var` calculates the variance of an input `x`.
- `sd` calculates the standard deviation of an input `x`.
- `IQR` calculates the interquartile range of an input `x`.
- `min` calculates the minimum value of an input `x`. 
- `max` calculates the maximum value of an input `x`. 
- `range` returns a vector containing the minimum and maximum of all given arguments.  

### Factor Input 

For factor input, we can extract counts and percentages as summary by using `table`. 

Here I create an example dataset with colors 20 that are `Yellow`, 10 that are `Green` and 50 that are `Blue`.
```{r}
(my_colors = as.factor(c(rep("Yellow", 20), rep("Green", 10), rep("Blue", 50))))
```

By using `table` we can see a frequency (count) table of the colors. 

```{r}
table(my_colors)
```

By diving the table by the number of total colors, we can get percentages instead. 

```{r}
table(my_colors) / length(my_colors)
```

### Dataset 
