## Array {#array}

Simply, vectors are one-dimensional and matrices are two-dimensional objects, and these homogenous objects are also known as **arrays**. We can extend the array into more than two dimensions by specifying the `dim` parameter within the `array` function. Arrays are rarer than matrices and vectors, but worth being aware of.

Below we initialize a three dimensional array, from 2 to 13. 
```{r}
three_dim <- array(data = 2:13, dim = c(2, 3, 2))
three_dim
```

As you can see, two 2x3 matrices were created containing numbers from 2 to 13. The `, , *` syntax describes which specific dimension the matrix is contained in. 
Like what we experimented with matrices, we can extract and manipulate information. 

```{r}
length(three_dim) # length is 12 
dim(three_dim) # dimensions are 2x3x2
is.array(three_dim) # yes it is an array!
```

Note that when changing the dimension names, we need to save it as a `list`. We will talk more about lists in the coming section. 

```{r}
dimnames(three_dim) <- list(c("apple", "banana"), c("one", "two", "three"), c("Walmart", "Costco")) # feel free to make 
three_dim
```


## List

A list is one of the commonly used heterogeneous data structures, which depicts a generic vector containing other object types. For example, we can have a numeric vector as one element, a matrix, and a character vector as another element. Here we can create a list that contains all these different element types. 

```{r}
num_vec <- c(160, 188, 140)
char_vec <- c("Height", "Weight", "Length")
my_mat <- matrix(0, nrow = 5, ncol = 5)

# List Initialization without Names 
# list(num_vec, char_vec, my_mat)

# List Initialization with Custom Names 
list(Number <- num_vec, Character = char_vec, Empty_Mat = my_mat)
```

Let's store this list in a variable. 

```{r}
my_list <- list(Number = num_vec, Character = char_vec, Empty_Mat = my_mat)
```

We can subset lists like we did previously. 

```{r}
# Extract the first and third element 
my_list[c(1, 3)]

# Compare these two subsets  
my_list[1]
my_list[[1]]
```

Do you see what the difference is? One generally uses `[[` to select any single element, whereas `[` returns a list of the selected elements. Using one bracket retains the class information, while using two brackets simplifies the list into a numeric vector. 

We can further subset the list. 

```{r}
# First element of the $Number list 
my_list[[1]][1]

# Why didn't this work? 
my_list[1][1]
```

