# (PART) Introduction to R {-}

# Data Structures {#data}

There are different data types that are commonly used in R among which the most important ones are the following:

- **Numeric** (or double): these are used to store real numbers. Examples: -4, 12.4532, 6.

- **Integer**: examples: 2L, 12L.

- **Logical** (or boolean): examples: `TRUE`, `FALSE`.

- **Character**: examples: `"a"`, `"Bonjour"`. 

In R there are basically five types of data structures in which elements can be stored. A data structure is said to *homogeneous* if it only contains elements of the same type (for example it only contains character or only numeric values) and *heterogenous* if it contains elements of more than one type. The five types of data structrures are commonly summarized in a table similar to the one below [see e.g. @wickham2014advanced]:

Table: (\#tab:ds) Five most common types of data structrures used in R [@wickham2014advanced].

Dimension      Homogenous    Heterogeneous
------------- ------------- ----------------
1               Vector        List
2               Matrix        Data Frame
n               Array

Consider a simple data set of the top five male single tennis players presented below:

Table: (\#tab:tennis) Five best male single tennis players as ranked by ATP  (07-15-2017).

 Name              Date of Birth         Born                Country            ATP Ranking     Prize Money    Win Percentage   Grand Slam Wins
----------------  ---------------  ---------------------- -------------------  ---------------  -------------- ---------------- ------------------
Andy Murray        15 May 1987     Glasgow, Scotland       Great Britain            1             60,449,649         78.07               9
Rafael Nadal       3 June 1986     Manacor, Spain           Spain                   2            85,920,132         82.48               15
Stan Wawrinka      28 March 1985   Lausanne, Switzerland    Switzerland             3            30,577,981         63.96              5
Novak Djokovic    22 May 1987       Belgrade, Serbia          Serbia                4            109,447,408        82.77              12
Roger Federer      8 August 1981    Basel, Switzerland      Switzerland            5            104,445,185        81.80              18

Notice that this data set contains a variety of data types; in the next sections we will use these data to illustrate how to work with five common data structures.

## Vectors 

A vector has three important properties:

```{block2, note-text, type='rmdimportant'}
- The **Type** of objects contained in the vector. The function `typeof()` returns a description of the type of objects in a vector.
- The **Length** of a vector indicates the number of elements in the vector. This information can be obtained using the function `length()`.
- **Attributes** are metadata attached to a vector. The functions `attr()` and `attributes()` can be used to store and retrive attributes (more details can be found in Section \@ref(vectattr))
```

`c()` is a generic function which combines arguments to form a vector. All arguments are coerced to a common type which is the type of the returned value, and all attributes except names are removed. For example, consider the number of grand slams won by the five players considered in the eighth column of Table \@ref(tab:tennis): 

```{r}
grand_slam_win <- c(9, 15, 5, 12, 18)
```

To display the values stored in `grand_slam_win` we could simply enter the following in the R console:

```{r}
grand_slam_win
```

Alternatively, we could have created and displayed the value by using `()` around the definition of the object itself as follows:

```{r}
(grand_slam_win <- c(9, 15, 5, 12, 18))
```

Various forms of "nested concatenation" can be used to defined vectors, for example we could also define `grand_slam_win` as

```{r}
(grand_slam_win <- c(9, c(15, 5, c(12, c(18)))))
```

This approach is often used to assemble vectors in various ways.

It is also possible to define vector with characters, for example we could define a vector with the player names as follows:

```{r}
(players <- c("Andy Murray", "Rafael Nadal", "Stan Wawrinka", 
             "Novak Djokovic", "Roger Federer"))
```

### Type 

We can evaluate the kind or type of elements that are stored in a vector using the function `typeof()`. For example, for the vectors we just created we obtain:

```{r}
typeof(grand_slam_win)
typeof(players)
```

This is a little surprising as all the elements in `grand_slam_win` are integers and it would therefore seem natural to expect this as an output of the function `typeof()`. This is because R considers any number as a "double" by default, except when adding the suffix `L` after an integer. For example:

```{r}
typeof(1)
typeof(1L)
```

Therefore, we could express `grand_slam_win` as follows:

```{r}
(grand_slam_win_int <- c(9L, 15L, 5L, 12L, 18L))
typeof(grand_slam_win_int)
```

In general, the difference between the two is relatively unimportant.

### Coercion

As indicated earlier, a vector has a homogeneous data structure meaning that it can only contain a single type among all the data types. Therefore, when more than one data type is provided, R *coerces* the data into a "shared" type. To identify this "shared" type we can use this simple rule:

\begin{equation*}
 \text{logical} < \text{integer} < \text{numeric} < \text{character},
\end{equation*}

which simply means that if a vector contains more than one data type, the "shared" type will be that of the "largest" type according to the progression shown above. For example:

```{r}
# Logical + numeric
(mix_logic_int <- c(TRUE, 12, 0.5))
typeof(mix_logic_int)

# Numeric + character
(mix_int_char <- c(14.3, "Hi"))
typeof(mix_int_char)
```


### Subsetting

Naturally, it is possible to "subset" the values of in our vector in many ways. Essentially, there are four main ways of subsetting a vector. Here we'll only discuss the first three:

- **Positive Index**: We can *access* or *subset* the $i$-th element of a vector by simply using `grand_slam_win[i]` where $i$ is a positive number between 1 and length of the vector.

```{r}
# Accesing the first element
grand_slam_win[1]

# Accesing the third and first value
grand_slam_win[c(3, 1)]

# Duplicated indices yield duplicated values
grand_slam_win[c(1, 1, 2, 2, 3, 4)]
```

- **Negative Index**: We *remove* elements in a vector using negative indices:

```{r}
# Removing the second obervation
grand_slam_win[-2]

# Removing the first and fourth obserations
grand_slam_win[c(-1, -4)]
```

- **Logical Indices**: Another useful approach is based on *logical* operators:

```{r}
# Access the first and fourth observations
grand_slam_win[c(TRUE, FALSE, FALSE, TRUE, FALSE)]
```

```{block2, type='rmdcaution'}
Note that it is not permitted to "mix" positive and negative indices. For example, `grand_slam_win[c(-1, 2)]` would produce an error message.
```

### Attributes {#vectattr}

Let's suppose that we conducted an experiment under specific conditions, say a date and a place which should be stored as attributes of the object containing the results of this experiment. Indeed, objects can have arbitrary additional attributes that are used to store metadata on the object of interest. For example:

```{r}
attr(grand_slam_win, "date") <- "07-15-2017"
attr(grand_slam_win, "type") <- "Men, Singles"
```

To display the vector with its attributes

```{r}
grand_slam_win
```

To only display the attributes we can use

```{r}
attributes(grand_slam_win)
```

It is also possible to extract a specific attribute

```{r}
attr(grand_slam_win, "date")
```

### Adding Labels {#addlab}

In some cases, it is useful to augment vector elements with labels. For example, we could define the vector `grand_slam_win` and associate the name of each corresponding athlete as a labels, i.e.

```{r}
(grand_slam_win <- c("Andy Murray" = 9, "Rafael Nadal" = 15, 
                   "Stan Wawrinka" = 5, "Novak Djokovic" = 12, 
                   "Roger Federer" = 18))
```

The main advantage of this approach is that the number of grand slams won can now be referred to by the player's name. For example:

```{r}
grand_slam_win["Andy Murray"]
grand_slam_win[c("Andy Murray","Roger Federer")]
```

All labels (athlete names in this case) can be obtained with the function `names()`, i.e.

```{r}
names(grand_slam_win)
```

### Working with Dates

When working with dates it is useful to treat them as actual dates rather than character strings that *look* like dates (to a human) but don't otherwise *behave* like dates. For example, consider a vector of three dates: `c("03-21-2015", "12-13-2011", "06-27-2008")`. 

The `sort()` function returns the elements of a vector in ascending order, but since these dates are actually just character strings that *look* like dates (to a human), R sorts them in alphanumeric order (for characters) rather than chronological order (for dates):

```{r}
# The `sort()` function sorts elements in a vector in ascending order
sort(c("03-21-2015", "12-31-2011", "06-27-2008", "01-01-2012"))
```

Converting the character strings to "yyyy-mm-dd" would solve our sorting problem, but perhaps we also want to calculate the number of days between two events that are several months or years apart. 

The `as.Date()` function is one straight-forward method for converting character strings into dates that can be used as such.  The typical syntax is of the form:

```{r, eval = FALSE}
as.Dates(<vector of dates>, format = <your format>)
```

Considering the dates of birth presented in Table \@ref(tab:tennis) we can save them in an appropriate format using:

```{r}
(players_dob <- as.Date(c("15 May 1987", "3 June 1986", "28 March 1985", 
                         "22 May 1987", "8 August 1981"), 
                       format = "%d %b %Y"))
```

Note the syntax of `format = "%d %b %Y"`.  The following table shows common format elements for use with the `as.Date()` function: 

Table: (\#tab:dateFormats) Common date formatting elements for use with `as.Date()` reproduced from [http://www.statmethods.net/](http://www.statmethods.net/input/dates.html).

Symbol   Meaning                    Example 
------- -------------------------- -------------
%d        day as a number (0-31)     01-31
%a        abbreviated weekday        Mon
%A	      unabbreviated weekday      Monday
%m        month (00-12)	             00-12
%b        abbreviated month          Jan
%B        unabbreviated month	       January
%y        2-digit year               07
%Y        4-digit year               2007


There are many advantages to recommend the `as.Date()` format (in addition to proper sorting). For example, the subtraction between two dates becomes more meaningful and yields to the difference in days between them. As an example, the number of days between Rafael Nadal's and Andy Murray's dates of birth can be obtained as

```{r}
players_dob[1] - players_dob[2]
```

In addition, subsetting becomes also more intuitive and to find out the players born after 1 January 1986 we can simply run: 

```{r}
players[players_dob > "1986-01-01"]
```

There are many other reason for using this format (or other date formats). A more detailed discussion on this topic can for example be found in [Cole Beck's notes](http://biostat.mc.vanderbilt.edu/wiki/pub/Main/ColeBeck/datestimes.pdf).

### Useful Functions with Vectors

The reason for extracting or creating vectors often lies in the need to collect information from them. For this purpose, a series of useful functions are available that allow to extract information or arrange the vector elements in a certain manner which can be of interest to the user. Among the most commonly used functions we can find the following ones 

`length()` `sum()` `mean()` `order()` and `sort()` 

whose name is self-explanatory in most cases. For example we have

```{r}
length(grand_slam_win)
sum(grand_slam_win)
mean(grand_slam_win)
```

To sort the players by number of grand slam wins, we could use the function `order()` which returns the *position* of the elements of a vector sorted in an ascending order,

```{r}
order(grand_slam_win)
```

Therefore, we can sort the players in ascending order of wins as follows

```{r}
players[order(grand_slam_win)]
```

which implies that Roger Federer won most grand slams. Another related function is `sort()` which simply sorts the elements of a vector in an ascending manner. For example,

```{r}
sort(grand_slam_win)
```

which is compact version of

```{r}
grand_slam_win[order(grand_slam_win)]
```

It is also possible to use the functions `sort()` and `order()` with characters and dates. For example, to sort the players' names alphabetically (by first name) we can use:

```{r}
sort(players)
```

Similarly, we can sort players by age (oldest first)

```{r}
players[order(players_dob)]
```

or in an reversed manner (oldest last):

```{r}
players[order(players_dob, decreasing = TRUE)]
```

There are of course many other useful functions that allow to deal with vectors which we will not mention in this section but can be found in a variety of references [see e.g. @wickham2014advanced].


### Creating sequences

When using R for statistical programming and data analysis it is very common to create sequences of numbers. Here are three common ways used to create such sequences:

- `from:to`: This method is quite intuitive and very compact. For example:

```{r}
1:3
(x <- 3:1)
(y <- -1:-4)
(z <- 1.3:3)
```

- `seq_len(n)`: This function provides a simple way to generate a sequence from 1 to an arbitrary number `n`. In general, `1:n` and `seq_len(n)` are equivalent with the notable exceptions where `n = 0` and `n < 0`. The reason for these exceptions will become clear in Section \@ref(forloop). Let's see a few examples:

```{r}
n <- 3
1:n
seq_len(n)

n <- 0
1:n
seq_len(n)
```

- `seq(a, b, by/length.out = d)`: This function can be used to create more "complex" sequences. It can either be used to create a sequence from `a` to `b` by increments of `d` (using the option `by`) or of a total length of `d` (using the option `length.out`). A few examples:

```{r}
(x <- seq(1, 2.8, by = 0.4))
(y <- seq(1, 2.8, length.out = 6))
```

This can be combined with the `rep()` function to create vectors with repeated values or sequences, for example:

```{r}
rep(c(1,2), times = 3, each = 1)
rep(c(1,2), times = 1, each = 3)
```

where the option `times` allows to specify how many times the object needs to be repeated and `each` regulates how many times each element in the object is repeated.

It is also possible to generates sequences of dates using the function `seq()`. For example, to generate a sequence of 10 dates between the dates of birth of Andy Murray and Rafael Nadal we can use

```{r}
seq(from = players_dob[1], to = players_dob[2], length.out = 10)
```

Similarly, we can create a sequence between the two dates by increment of one week (backward)

```{r}
seq(players_dob[1], players_dob[2], by = "-1 week")
```

or by increment of one month (forward)

```{r}
seq(players_dob[2], players_dob[1], by = "1 month")
```

### Example: Apple Stock Price 

Suppose that one is interested in analyzing the behavior of Apple's stock price over the last three months. The first thing that is needed to perform such analysis is to recover (automatically) today's date.  In R, this can be obtained as follows

```{r}
(today <- Sys.Date())
```

Once this is done, we can obtain the date which is exactly three months ago by using

```{r}
(three_months_ago <- seq(today, length = 2, by = "-3 months")[2])
```

With this information, we can now download Apple's stock price and represent these stocks through a candlestick chart which summarizes information on daily opening and closing prices as well as minimum and maximum prices. These charts are often used with the hope of detecting trading patterns over a given period of time.

```{r candleAAPL, message = FALSE, fig.height = 5, fig.width = 6, fig.align = "center", warning = FALSE, results = 'hide', fig.cap = "Candlestick chart for Apple's stock price for the last three months."}
library(quantmod)
getSymbols("AAPL", from = three_months_ago, to = today)
candleChart(AAPL, theme='white', type='candles')
```

Using the price contained in the object we downloaded (i.e. `AAPL`), we can compute Apple's arithmetic returns which are defined as follows

\begin{equation}
  r_t = \frac{S_t - S_{t-1}}{S_{t-1}},
\end{equation}

where $r_t$ are the returns at time *t* and $S_t$ is the stock price. This is implemented in the function `ClCl()` within the `quantmod` package. For example, we can create a vector of returns as follows 

```{r}
AAPL_returns <- na.omit(ClCl(AAPL))
```

where `na.omit()` is used to remove missing values in the stock prices vector since, if we have $n+1$ stock prices, we will only have $n$ returns and therefore the first return cannot be computed. We can now compute the mean and median of the returns over the considered period.

```{r}
mean(AAPL_returns)
median(AAPL_returns)
```

However, a statistic that is of particular interest to financial operators is the Excess Kurtosis which, for a random variable that we denote as $X$, can be defined as

\begin{equation}
  \text{Kurt} = \frac{{\mathbb{E}}\left[\left(X - \mathbb{E}[X]\right)^4\right]}{\left({\mathbb{E}}\left[\left(X - \mathbb{E}[X]\right)^2\right]\right)^2} - 3.
\end{equation}

The reason for defining this statistic as *Excess* Kurtosis lies in the fact that the standardized kurtosis is compared to that of a Gaussian distribution (whose kurtosis is equal to 3) which has exponentially decaying tails. Consequently, if the Excess Kurtosis is positive, this implies that the distribution has heavier tails than a Gaussian and therefore has higher probabilities of extreme events occurring. Given this statistic, it is useful to compute this on the observed data and for this purpose a common estimator of the excess Kurtosis is

\begin{equation}
  k = \frac{\frac{1}{n} \sum_{t = 1}^{n} \left(r_t -\bar{r}\right)^4}{\left(\frac{1}{n} \sum_{t = 1}^{n} \left(r_t -\bar{r}\right)^2 \right)^2} - 3,
\end{equation}

where $\bar{k}$ denotes the sample average of the returns, i.e.

\begin{equation}
  \bar{k} = \frac{1}{n} \sum_{i = 1}^n r_i.
\end{equation}

In R, this can simply be done as follows:

```{r}
mu <- mean(AAPL_returns)
(k <- mean((AAPL_returns - mu)^4)/(mean((AAPL_returns - mu)^2))^2 - 3)
```
 
Therefore, we observe an estimated Excess Kurtosis of `r round(k,2)` which is quite high tends to indicate the returns have a heavier tails than the normal distribution. In Chapter \@ref(#control), we will revisit this example and investigate if there is **enough evidence** to conclude that Apple's stock has Excess Kurtosis larger than zero.
 
## Matrix

Matrices are another extremely common data structure in R. Compared to vectors, matrices have an additional dimension which, for example, allows to stock multiple equidimensional vectors within the same object. Below is an example of how to create a matrix in R:

```{r}
(mat <- matrix(1:12, ncol = 4,  nrow = 3))
```

Notice that the first argument to the function is a vector (in this case a vector with increasing elements from 1 to 12) which is then transformed into a matrix with four columns (`ncol = 4`) and three rows (`nrow = 3`). 

```{block2, type='rmdtip'}

By default, the vectors are transformed into matrices by placing the elements by column (i.e. starting from the top of the first column to the bottom and then passing to the following column until all columns are full). If you wish to fill the matrix by row, all you need to do is specify the argument `byrow = T`.

```

```{block2, type='rmdwarning'}

Usually the length of the vector (i.e. number of elements in the vector) is the result of the multiplication between the number of columns and number of rows. What happens if the vector has fewer elements for the same matrix dimension? What happens if the vector has more elements? 

```

It is often the case however that we already have equidimensional vectors available and we wish to stock them into a matrix. In these cases, two useful functions are `cbind()` and `rbind()` where the first function stocks the vectors vertically side-by-side while the second stocks the vectors horizontally one below the other. An example of the former is given below:

```{r}
players <- c("Andy Murray", "Rafael Nadal", "Stan Wawrinka", 
             "Novak Djokovic", "Roger Federer")
grand_slam_win <- c(9, 15, 5, 12, 18)
win_percentage <- c(78.07, 82.48, 63.96, 82.77, 81.80)
(mat <- cbind(grand_slam_win, win_percentage))
```

The result in this case is a $5 \times 2$ matrix (while with `rbind()` it would have been a $2 \times 5$ matrix). Once the matrix is defined, we can assign names to its rows and columns by using respectively `rownames` and `colnames`. Of course, the number of names must correspond to the respective matrix dimensions as shown in the following example where each row corresponds to a specific player (thereby using the `players` vector) and each column corresponds to a specific statistic on the players.

```{r}
rownames(mat) <- players
colnames(mat) <- c("GS win", "Win rate")
mat
```

### Subsetting

As for vectors, it is possible to subset the elements of a matrix. However, in the case of matrices we're dealing with two-dimensional data structures and it is therefore necessary to specify the position of the elements of interest in both dimensions. For this purpose, as with vectors, we can use `[ ]` but, as opposed to vectors, we need to add `,` within the square brackets where the rows are specified before the comma and the columns after it. Below are a few examples:

```{r}
mat[c("Stan Wawrinka", "Roger Federer"), ]
mat[c(1, 3), ]
mat[, 2]
mat[1:3, 1]
```

It can be noticed that, when a space is left blank before or after the comma, this means that respectively all the rows or all the columns are considered.

### Linear Algebra 

As with vectors, there are some useful functions that can be used with matrices. A first example is the function `dim()` that allows to determine the dimension of a matrix. For example, consider the following $4 \times 2$ matrix

\begin{equation*}
\mathbf{A} = \left[
\begin{matrix}
1 & 5\\
2 & 6\\
3 & 7\\
4 & 8
\end{matrix}
\right]
\end{equation*}

which can be created in R as follows:
```{r}
(A <- matrix(1:8, 4, 2))
```

Therefore, we expect `dim(A)` to return the vector `c(4, 2)`. Indeed, we have

```{r}
dim(A)
```

Next, we consider the function `t()` allows transpose a matrix. For example, $\mathbf{A}^T$ is equal to:

\begin{equation*}
\mathbf{A}^T = \left[
\begin{matrix}
1 & 2 & 3 & 4\\
5 & 6 & 7 & 8
\end{matrix}
\right],
\end{equation*}

which is a $2 \times 4$ matrix. In R, we achieve this as follows

```{r}
(At <- t(A))
dim(At)
```

Aside from playing with matrix dimensions, matrix algebraic operations have specific commands. For example, the operator `%*%` is used in R to denote matrix multiplication while, as opposed to scalar objects, the regular product operator `*` performs the Hadamard product (or element by element product) when applied to matrices. For example, consider the following matrix product

\begin{equation*}
  \mathbf{B} = \mathbf{A}^T \mathbf{A} =   \left[
\begin{matrix}
30 & 70\\
70 & 174
\end{matrix}
\right],
\end{equation*}

which can be done in R as follows:

```{r}
(B <- At %*% A)
```

Other common matrix operations include finding the determinant of a matrix and finding its inverse. These are often used, for example, when computing the likelihood function for a variable following a Gaussian distribution or when simulating time series or spatial data. The functions that perform these operations are `det()` and `solve()` that respectively find the determinant and the inverse of a matrix (which necessarily has to be square). The function `det()` can be used to compute the determinant of a (squared) matrix. In the case of a $2 \times 2$ matrix, there exists a simple solution for the determinant which is

\begin{equation*}
\text{det} \left( \mathbf{D} \right) = \text{det} \left( \left[
\begin{matrix}
d_1 & d_2\\
d_3 & d_4
\end{matrix}
\right] \right) = d_1 d_4 - d_2 d_3.
\end{equation*}

Consider the matrix $\mathbf{B}$, we have

\begin{equation*}
  \text{det} \left( \mathbf{B}\right) = 30 \cdot 174 - 70^2 = 320.
\end{equation*}

In R, we can simply do

```{r}
det(B)
```

The function `solve()` is also an important function when working with matrices as it allows to inverse a matrix. It is worth remembering that a square matrix that is not invertible (i.e. $\mathbf{A}^{-1}$ doesn't exist) is called *singular* and the determinant offers a way to "check" if this is the case for a given matrix. Indeed, a square matrix is singular if and only if its determinant is 0. Therefore, in the case of $\mathbf{B}$, we should be able to compute its inverse. As for the determinant, there exists a formula to compute the inverse of $2 \times 2$ matrices, i.e.

\begin{equation*}
 \mathbf{D}^{-1} = \left[
\begin{matrix}
d_1 & d_2\\
d_3 & d_4
\end{matrix}
\right]^{-1} = \frac{1}{\text{det}\left( \mathbf{D} \right)} \left[
\begin{matrix}
\phantom{-}d_4 & -d_2\\
-d_3 & \phantom{-}d_1
\end{matrix}
\right].
\end{equation*}

Considering the matrix $\mathbf{B}$, we obtain

\begin{equation*}
 \mathbf{B}^{-1} = \left[
\begin{matrix}
30 & 70\\
70 & 174
\end{matrix}
\right]^{-1} = \frac{1}{320}\left[
\begin{matrix}
  \phantom{-}174 & -70\\
-70 & \phantom{-}30
\end{matrix}
\right] = 
\end{equation*}

```{r}
(B_inv <- solve(B))
```

Finally, we can verify that 

\begin{equation*}
\mathbf{G} = \mathbf{B} \mathbf{B}^{-1},
\end{equation*}

should be equal to the identity matrix,

```{r}
(G <- B %*% B_inv)
```

The result is of course extremely close but $\mathbf{G}$ is not exactly equal to the identity matrix due to rounding and other numerical errors.

Another function of interest is the function `diag()` that can be used to extract the diagonal of a matrix. For example, we have

\begin{equation*}
\text{diag} \left( \mathbf{B} \right) = \left[30 \;\; 174\right],
\end{equation*}

which can be done in R as follows:

```{r}
diag(B)
```

Therefore, the function `diag()` allows to easily compute the trace of matrix (i.e. the sum of the diagonal elements). For example,

\begin{equation*}
\text{tr} \left( \mathbf{B} \right) = 204,
\end{equation*}

or in R

```{r}
sum(diag(B))
```

Another use of the function `diag()` is to create diagonal matrices. Indeed, if the argument of this function is a vector, its behavior is the following:

\begin{equation*}
  \text{diag} \left(\left[a_1 \;\; a_2 \;\; \cdots \;\; a_n\right]\right) = \left[
\begin{matrix}
a_1     & 0       & \cdots & 0  \\
0       & a_2     & \cdots & 0  \\
\vdots  & \vdots  & \ddots       & \vdots    \\
0       & 0       &   \cdots     & a_n
\end{matrix}
\right].
\end{equation*}

Therefore, this provides a simple way of creating an identity matrix by combining the functions `diag()` and `rep()` (discussed in the previous section) as follows:

```{r}
n = 4
(ident <- diag(rep(1, n)))
```

### Example: Summary Statistics with Matrix Notation

A simple example of the operations we discussed in the previous section is given by many common statistics that can be re-expressed using matrix notation. As an example, we will consider three common statistics that are the sample mean, variance and covariance. Let us consider the following two samples of size $n$

\begin{equation*}
  \begin{aligned}
    \mathbf{x} &= \left[x_1 \;\; x_2 \; \;\cdots \;\; x_n\right]^T\\
    \mathbf{y} &= \left[y_1 \;\;\; y_2 \; \;\;\cdots \;\;\; y_n\right]^T.
  \end{aligned}
\end{equation*}

The sample mean of $\mathbf{x}$ is

\begin{equation*}
  \bar{x} = \frac{1}{n} \sum_{i = 1}^{n} x_i,
\end{equation*}

and its sample variance is

\begin{equation*}
  s_x^2 = \frac{1}{n} \sum_{i = 1}^n \left(x_i - \bar{x}\right)^2.
\end{equation*}

The sample covariance between $\mathbf{x}$ and $\mathbf{y}$ is

\begin{equation*}
  s_{x,y} = \frac{1}{n} \sum_{i = 1}^n \left(X_i - \bar{x}\right) \left(Y_i - \bar{y}\right),
\end{equation*}

where $\bar{y}$ denotes the sample mean of $\mathbf{y}$.

Consider the sample mean, this statistic can be expressed in matrix notation as follows

\begin{equation*}
  \bar{x} = \frac{1}{n} \sum_{i = 1}^{n} x_i =  \frac{1}{n} \mathbf{x}^T \mathbf{1},
\end{equation*}

where $\mathbf{1}$ is a column vector of $n$ ones.

\begin{equation*}
  \begin{aligned}
    s_x^2 &= \frac{1}{n} \sum_{i = 1}^n \left(x_i - \bar{x}\right)^2 = \frac{1}{n} \sum_{i = 1}^n x_i^2 - \bar{x}^2 = \frac{1}{n} \mathbf{x}^T \mathbf{x} - \bar{x}^2\\
    &= \frac{1}{n} \mathbf{x}^T \mathbf{x} - \left(\frac{1}{n} \mathbf{x}^T \mathbf{1}\right)^2 = \frac{1}{n} \left(\mathbf{x}^T \mathbf{x} - \frac{1}{n} \mathbf{x}^T \mathbf{1} \mathbf{1}^T \mathbf{x}\right)\\
    &= \frac{1}{n}\mathbf{x}^T \left( \mathbf{I} - \frac{1}{n} \mathbf{1} \mathbf{1}^T \right) \mathbf{x} = \frac{1}{n}\mathbf{x}^T \mathbf{H} \mathbf{x},
  \end{aligned}
\end{equation*}

where $\mathbf{H} = \mathbf{I} - \frac{1}{n} \mathbf{1} \mathbf{1}^T$. This matrix is often called the *centering* matrix. Similarly, for the sample covariance we obtain

\begin{equation*}
  \begin{aligned}
    s_{x,y} &= \frac{1}{n} \sum_{i = 1}^n \left(x_i - \bar{x}\right) \left(y_i - \bar{y}\right) = \frac{1}{n}\mathbf{x}^T \mathbf{H} \mathbf{y}.
  \end{aligned}
\end{equation*}

In the code below we verify the validity of these results by comparing the value of the three statistics based on the different formulas.

```{r}
# Sample size
n <- 100

# Simulate random numbers from a zero mean normal distribution with
# variance equal to 4.
x <- rnorm(n, 0, sqrt(4))

# Simulate random numbers from normal distribution with mean 3 and
# variance equal to 1.
y <- rnorm(n, 3, 1)

# Note that x and y are independent.

# Sample mean
one <- rep(1, n)
(x_bar <- 1/n*sum(x))
(x_bar_mat <- 1/n*t(x)%*%one)

# Sample variance
H <- diag(rep(1, n)) - 1/n * one %*% t(one)
(s_x <- 1/n * sum((x - x_bar)^2))
(s_x_mat <- 1/n*t(x) %*% H %*% x)

# Sample covariance
y_bar <- 1/n*sum(y)
(s_xy <- 1/n*sum((x - x_bar)*(y - y_bar)))
(s_xy_mat <- 1/n*t(x) %*% H %*% y)
```

## Array {#array}

Simply, vectors are one-dimensional and matrices are two-dimensional objects, and these homogenous objects are also known as **arrays**. We can extend the array into more than two dimensions by specifying the `dim` parameter within the `array` function. Arrays are rarer than matrices and vectors, but worth being aware of.

Below we initialize a three dimensional array, from 2 to 13. 
```{r}
three_dim <- array(data = 2:13, dim = c(2, 3, 2))
three_dim
```

As you can see, two 2x3 matrices were created containing numbers from 2 to 13. The `, , *` syntax describes which specific dimension the matrix is contained in. 
Like what we experimented with matrices, we can extract and manipulate information. 

```{r}
length(three_dim) # length is 12 
dim(three_dim) # dimensions are 2x3x2
is.array(three_dim) # yes it is an array!
```

Note that when changing the dimension names, we need to save it as a `list`. We will talk more about lists in the coming section. 

```{r}
dimnames(three_dim) <- list(c("apple", "banana"), c("one", "two", "three"), c("Walmart", "Costco")) # feel free to make 
three_dim
```


## List

A list is one of the commonly used heterogeneous data structures, which depicts a generic vector containing other object types. For example, we can have a numeric vector as one element, a matrix, and a character vector as another element. Here we can create a list that contains all these different element types. 

```{r}
num_vec <- c(160, 188, 140)
char_vec <- c("Height", "Weight", "Length")
my_mat <- matrix(0, nrow = 5, ncol = 5)

# List Initialization without Names 
# list(num_vec, char_vec, my_mat)

# List Initialization with Custom Names 
list(Number <- num_vec, Character = char_vec, Empty_Mat = my_mat)
```

Let's store this list in a variable. 

```{r}
my_list <- list(Number = num_vec, Character = char_vec, Empty_Mat = my_mat)
```

We can subset lists like we did previously. 

```{r}
# Extract the first and third element 
my_list[c(1, 3)]

# Compare these two subsets  
my_list[1]
my_list[[1]]
```

Do you see what the difference is? One generally uses `[[` to select any single element, whereas `[` returns a list of the selected elements. Using one bracket retains the class information, while using two brackets simplifies the list into a numeric vector. 

We can further subset the list. 

```{r}
# First element of the $Number list 
my_list[[1]][1]

# Why didn't this work? 
my_list[1][1]
```


## Dataframe

*Needs some rewording* 
 
A data frame is used for storing data tables. A data frame is the most common way of storing data in R, it has a 2D structure and shares properties of both the matrix and the list. The table contains lists of equal-length vectors, and most datasets will have a data frame format. 

We can create a data frame using data.frame() 

```{r}
### Creation

players <- c("Andy Murray", "Rafael Nadal", "Stan Wawrinka", 
             "Novak Djokovic", "Roger Federer")

grand_slam_win <- c(9, 15, 5, 12, 18)

date_of_birth <- c("15 May 1987", "3 June 1986", "28 March 1985", 
                  "22 May 1981", "8 August 1981")

country = c("Great Britain", "Spain", "Switzerland", 
            "Serbia", "Switzerland")
ATP_ranking <- c(1, 2, 3, 4, 5)

prize_money <- c(60449649, 85920132, 30577981, 109447408, 104445185)

tennis <- data.frame(date_of_birth, grand_slam_win, country, 
                    ATP_ranking, prize_money)

dimnames(tennis)[[1]] <- players
tennis

```

We can check if we have achived our goal by using:

```{r}
is.data.frame(tennis)
```

### Combination

Different data frames can also be combined. Let say we want to add some ifomrmation to our initial table e.g. the player's height and if he is right-handed or letf-handed. 

We can do so by using `cbind()` and `rbind()`:

```{r}
height <- c(1.90, 1.85, 1.83, 1.88, 1.85)
hand <- c("R","L","R","R","R")

(tennis <- cbind(tennis, data.frame(height, hand)))
```

### Subsetting

Like for vectors, it is also possible to subset the values that we have stored in our data frames. Since data frames possess the characteristics of both lists and matrices: if you subset with a single vector, they behave like lists; if you subset with two vectors, they behave like matrices.

```{r}
# Let say we want only want to know the country and date of 
# birth of the players

# There are two ways to select columns from a data frame
# Like a list:
tennis[c("country", "date_of_birth")]

# Like a matrix
tennis[, c("country", "date_of_birth")]

# To acces a single element, let say the date of birth, 
# you can also use:
tennis$date_of_birth
```

### Application: Non-parametric bootstrap

Suppose we ask 10 students how much time they work at home for their calculus class, we obtain the following results (in hour)

```{r}
student_work <- c(0, 0, 0, 0.25, 0.25, 0.75, 0.75, 1, 1.25, 4)
```

We can compute the mean time spent

```{r}
mean(student_work)
```

*ADD SOMETHING ON T TEST*

```{r}
t.test(student_work)$conf.int
```

We can see that our confidence interval includes a negative values which clearly isn't meaningful. Solution: (non-parametric) bootstrap which works as follows..... ADD SOMETHING

Here is a simple function to implement this approach:

```{r}
# Number of boostrap replications
B <- 500

# Compute the length of vector
n <- length(student_work)

# Confidence level
alpha <- 0.05

# Initialisation of 
boot_mean <- rep(NA, B)

for (i in 1:B){
  student_work_star <- student_work[sample(1:n, replace = TRUE)]
  boot_mean[i] <- mean(student_work_star)
}

quantile(boot_mean, c(alpha/2, 1 - alpha/2))
#hist(boot_mean, probability = TRUE)

```
